Clustering in a hyperbolic model of complex networks.

### 2019-03-13 clustering_infinite_model.tex
finished revision of sections 5 and 6 and Appendix D

As mentioned earlier, we plan to make use of the Campbell-Mecke formula for comparing the clustering coefficient and function of $\GPo$ with certain quantities associated with $\Ginf$. We will be considering the Poisson process $\Pcal$ to which we add one additinal point $(0,y)$ on the $y$-axis. In some computations the height $y$ will be fixed, but eventually we shall take it exponentially distributed with parameter $\alpha$, and independent of $\Pcal$.
We refer to $(0,y)$ as ``the typical point''.

To provide some intu\"{\i}tion for this definition and name, note that we can alternatively view $\Pcal$ as follows. 
We take a constant intensity Poisson process on $\eR$ corresponding to the $x$-coordinates, and to each point
we attach a random ``mark'', corresponding to the $y$-coordinate, where the marks are i.i.d.~exponentially distributed with parameter $\alpha$.

@@

If $p = (x,y) \in \eR\times[0,\infty)$ is a point, not necessarily part of the Poisson process, then we will write

$$ \mu(y) = \mu(p) := \mu(\BallPo{p}). $$
Integrating the intensity function of $\Pcal$ over $\BallPo{p}$ gives, using $\alpha>\frac{1}{2}$,

$$ \begin{array}{rcl} 
\mu(y) & = & \int_{\BallPo{p}} f(x',y') \, dx' \,d y'  
 = \int_0^\infty \int_{-e^{(y+y')/2}}^{e^{(y+y')/2}} \frac{\alpha \nu}{\pi} e^{-\alpha y'} \, dx' \, dy' \\
& = & \int_0^\infty 2 e^{(y+y')/2} \frac{\alpha \nu}{\pi} e^{-\alpha y'} \, dy' 
 = \frac{2\alpha \nu e^{y/2}}{\pi} \int_0^\infty e^{(\frac12-\alpha)y'} \, dy' \\
& = & \frac{2\alpha \nu e^{y/2}}{\pi(\alpha-\frac12)} = \xi e^{y/2}.
\end{array} $$

@@

Before considering clustering we briefly investigate the distribution of the degree of 
the typical point. 
For $p = (x,y) \in \eR\times[0,\infty)$ we define
\begin{equation}
	\rho(p,k) := \Prob{\Po(\mu(y)) = k},
\end{equation}

where $\Po(\lambda)$ denotes a Poisson random variable with expectation $\lambda$. 
We will often write $\rho(y,k)$ instead of $\rho(p,k)$.

@@

Using the transformation of variables $z = \xi e^{\frac{y}{2}}$ (so $dy = \frac{2}{z}dz$), we compute
\begin{align*}
	p_k
    &= \frac{1}{k!} \int_0^\infty \left(\xi e^{\frac{y}{2}}\right)^k 
    	e^{-\xi e^{\frac{y}{2}}} \alpha e^{-\alpha y} \, dy\\
    &= \frac{\alpha \xi^{2\alpha}}{k!} \int_0^\infty 
    	\left(\xi e^{\frac{y}{2}}\right)^{k - 2\alpha} e^{-\xi e^{\frac{y}{2}}}
        \, dy\\
    &= \frac{2\alpha \xi^{2\alpha}}{k!} \int_{\xi}^{\infty} 
    	z^{k -2\alpha-1} e^{-z} \, dz\\
    &= \frac{2\alpha \xi^{2\alpha}\Gamma^+(k - 2\alpha, \xi)}{k!},
\end{align*}
where we recall that $\Gamma$ denotes the Gamma-function and $\Gamma^{+}$ the upper incomplete Gamma-function.
Note that, unsurprisingly, this is identical to the expression Gugelmann et al.~\cite{gugelmann2012random} gave for the limiting degree distribution of $G(n;\alpha,\nu)$. Using Stirling's approximation to the gamma function, we find that

@@

Let the random variable $C$ denote the clustering coefficient of the typical point $(0,y)$, in the graph obtained from $\Ginf$ by adding $(0,y)$. We {\em define}
\[
	\gamma := \Ee C, \quad \gamma(k) := \Ee\left( C | D = k \right).
\]
(Where we take the expectation over both the Poisson point process $\Pcal$ and $y \isd \text{exp}(\alpha)$, independently of the Poisson process $\Pcal$.) We shall show shortly that these take on the values stated in Theorem~\ref{thm:maincc} and~\ref{thm:mainkfixed}.

For any fixed value $y_0>0$, the set of points inside $\BallPo{(0,y_0)}$ is a Poisson process with intensity $f \cdot 1_{\BallPo{(0,y_0)}}$. As $\mu(\BallPo{(0,y_0)}) = \mu(y_0) = \xi e^{y_0/2} < \infty$, this can be described alternatively by the experiment where we first pick $N \isd \Po( \mu(y_0) )$ and then we take $N$ i.i.d.~points in $\BallPo{(0,y_0)}$ according to the probability density $f \cdot 1_{\BallPo{(0,y_0)}} / \mu(y_0)$. (That is, the intensity function of the Poisson point process, but set to zero outside of $\BallPo{(0,y_0)}$ and re-normalized in such a way that it integrates to one.) Hence, if we condition on the event that $y$ takes on some fixed value $y_0$ and that there are exactly $k$ points of $\Pcal$ inside $\BallPo{(0,y_0)}$, then those $k$ points behave like $k$ i.i.d.~points in $\BallPo{(0,y_0)}$ chosen according to the mentioned re-normalized probability density function. This shows that, for every $k\geq 2$:
\[
	\Ee\left( C | D=k, y=y_0 \right) 
	= \frac{1}{{k\choose 2}} \Ee\left( \sum_{1\leq i < j \leq k} \1_{\{u_i\in \BallPo{u_j}\}} \right)
	= \Exp{\1_{\{u_1\in\BallPo{u_2}\}}},
\]
where $u_1,\dots, u_n$ are i.i.d.~points in $\BallPo{(0,y_0)}$ with the above mentioned density.
Note that this does not depend on the value of $k$. For notational convenience, we will write 
\[ 
	P(y_0) := \Exp{\1_{\{u_1\in\BallPo{u_2}\}}},
\]
with $u_1, u_2$ as above.

@@

Recall that $P(y_0)$ is the probability that $u_1 = (x_1,y_1), u_2 = (x_2,y_2)$ are neighbours in $\Ginf$, where
$u_1, u_2$ are i.i.d.~with probability density $f \cdot \1_{\BallPo{(0,y_0)}} / \mu(y_0)$.
In particular

$$ \begin{array}{rcl} 
\Pee( y_i > t ) 
& = & 
\frac{\nu\alpha}{\pi \mu(y_0)} \int_t^\infty \int_{-e^{(s+y_0)/2}}^{e^{(s+y_0)/2}} e^{-\alpha s}\, d x \, d s \\
& = & 
\frac{\nu\alpha}{\pi \mu(y_0)} \int_t^\infty 2 e^{(s+y_0)/2} \cdot e^{-\alpha s}\, d s \\
& = & 
\frac{2 \nu\alpha e^{y_0/2} }{\pi \xi e^{y_0/2} (\alpha-\frac12) }  \cdot e^{(\frac12-\alpha)t} \\
& = & 
 e^{(\frac12-\alpha)t},
\end{array} $$

\noindent
using that $\mu(y_0) = \xi e^{y_0/2} = \left(\frac{2\alpha\nu}{\pi(\alpha-\frac12)}\right) e^{y_0/2}$.
Thus, $y_1, y_2$ are exponentially distributed with parameter $\alpha-\frac12$.
Now note that, for each $t>0$, the probability density $f \cdot 1_{\BallPo{(0,y_0)}} / \mu(y_0)$
is constant on $[-e^{(t+y_0)/2}, e^{(t+y_0)/2}] \times \{t\}$ and it is 
vanishes on $(-\infty, -e^{(t+y_0)/2}) \times \{t\} \cup (e^{(t+y_0)/2},\infty)\times\{t\}$.

>>>>

As mentioned earlier, we plan to make use of the Campbell-Mecke formula for comparing the clustering coefficient and function of $\GPo$ with certain quantities associated with $\Ginf$. We will be considering the Poisson process $\Pcal$ to which we add one additional point $(0,y)$ on the $y$-axis. In some computations the height $y$ will be fixed, but eventually we shall take it exponentially distributed with parameter $\alpha$, and independent of $\Pcal$.
We refer to $(0,y)$ as ``the typical point''.

To provide some intuition for this definition and name, note that we can alternatively view $\Pcal$ as follows. 
We take a constant intensity Poisson process on $\eR$ corresponding to the $x$-coordinates, and to each point
we attach a random ``mark'', corresponding to the $y$-coordinate, where the marks are i.i.d.~exponentially distributed with parameter $\alpha$.

@@

If $p = (x,y) \in \eR\times[0,\infty)$ is a point, not necessarily part of the Poisson process, then we will write
\[ 
	\mu(y) = \mu(p) := \mu(\BallPo{p}).
\]
Integrating the intensity function of $\Pcal$ over $\BallPo{p}$ gives, using $\alpha>\frac{1}{2}$,
\begin{align*}
	\mu(y) &= \int_{\BallPo{p}} f(x',y') \, dx' \,d y'  
 		= \int_0^\infty \int_{-e^{(y+y')/2}}^{e^{(y+y')/2}} \frac{\alpha \nu}{\pi} e^{-\alpha y'} \, dx' \, dy' \\
	&= \int_0^\infty 2 e^{(y+y')/2} \frac{\alpha \nu}{\pi} e^{-\alpha y'} \, dy' 
 		= \frac{2\alpha \nu e^{y/2}}{\pi} \int_0^\infty e^{(\frac12-\alpha)y'} \, dy' \\
	&= \frac{2\alpha \nu e^{y/2}}{\pi(\alpha-\frac12)} = \xi e^{y/2}.
\end{align*}

@@

Before considering clustering we briefly investigate the distribution of the degree of the typical point. 
For $p = (x,y) \in \eR\times[0,\infty)$ we define
\begin{equation}\label{eq:def_degree_inf_model}
	\rho(p,k) := \Prob{\Po(\Mu{p}) = k},
\end{equation}
where $\Po(\lambda)$ denotes a Poisson random variable with expectation $\lambda$. We will often write $\rho(y,k)$ instead of $\rho(p,k)$.

@@

Using the transformation of variables $z = \xi e^{\frac{y}{2}}$ (so $dy = \frac{2}{z}dz$), we compute
\begin{align*}
	p_k
    &= \frac{1}{k!} \int_0^\infty \left(\xi e^{\frac{y}{2}}\right)^k 
    	e^{-\xi e^{\frac{y}{2}}} \alpha e^{-\alpha y} \, dy
    	= \frac{\alpha \xi^{2\alpha}}{k!} \int_0^\infty 
    	\left(\xi e^{\frac{y}{2}}\right)^{k - 2\alpha} e^{-\xi e^{\frac{y}{2}}}
        \, dy\\
    &= \frac{2\alpha \xi^{2\alpha}}{k!} \int_{\xi}^{\infty} 
    	z^{k -2\alpha-1} e^{-z} \, dz
    	= \frac{2\alpha \xi^{2\alpha}\Gamma^+(k - 2\alpha, \xi)}{k!},
\end{align*}
where we recall that $\Gamma$ denotes the Gamma-function and $\Gamma^{+}$ the upper incomplete Gamma-function.
Note that, unsurprisingly, this is identical to the expression Gugelmann et al.~\cite{gugelmann2012random} gave for the limiting degree distribution of $G(n;\alpha,\nu)$. Using Stirling's approximation to the gamma function, we find that

@@

Let the random variable $C$ denote the clustering coefficient of the typical point $(0,y)$, in the graph obtained from $\Ginf$ by adding $(0,y)$. We now {\em define}
\[
	\gamma := \Ee C, \quad \gamma(k) := \Ee\left( C | D = k \right).
\]
(Where we take the expectation over both the Poisson point process $\Pcal$ and $y \isd \text{exp}(\alpha)$, independently of the Poisson process $\Pcal$.) We shall show shortly that these take on the values stated in Theorem~\ref{thm:maincc} and~\ref{thm:mainkfixed}.

For any fixed value $y_0>0$, the set of points inside $\BallPo{(0,y_0)}$ is a Poisson process with intensity $f \cdot 1_{\BallPo{(0,y_0)}}$. As $\mu(\BallPo{(0,y_0)}) = \mu(y_0) = \xi e^{y_0/2} < \infty$, this can be described alternatively by first picking $N \isd \Po( \mu(y_0) )$ and then taking $N$ i.i.d.~points in $\BallPo{(0,y_0)}$ according to the probability density $f \cdot 1_{\BallPo{(0,y_0)}} / \mu(y_0)$. (That is, the intensity function of the Poisson point process, but set to zero outside of $\BallPo{(0,y_0)}$ and re-normalized in such a way that it integrates to one.) Hence, if we condition on the event that $y$ takes on some fixed value $y_0$ and that there are exactly $k$ points of $\Pcal$ inside $\BallPo{(0,y_0)}$, then those $k$ points behave like $k$ i.i.d.~points in $\BallPo{(0,y_0)}$ chosen according to the mentioned re-normalized probability density function. This shows that, for every $k\geq 2$:
\[
	\Ee\left( C | D=k, y=y_0 \right) 
	= \frac{1}{{k\choose 2}} \Ee\left( \sum_{1\leq i < j \leq k} \1_{\{u_i\in \BallPo{u_j}\}} \right)
	= \Exp{\1_{\{u_1\in\BallPo{u_2}\}}},
\]
where $u_1,\dots, u_n$ are i.i.d.~points in $\BallPo{(0,y_0)}$ with the above mentioned density.
Note that this does not depend on the value of $k$. For notational convenience, we will write 
\[ 
	P(y_0) := \Exp{\1_{\{u_1\in\BallPo{u_2}\}}},
\]
with $u_1, u_2$ as above.

@@

Recall that $P(y_0)$ is the probability that $u_1 = (x_1,y_1), u_2 = (x_2,y_2)$ are neighbours in $\Ginf$, where
$u_1, u_2$ are i.i.d.~with probability density $f \cdot \1_{\BallPo{(0,y_0)}} / \mu(y_0)$.
In particular
\begin{align*}
	\Prob{y_i > t} &= \frac{\nu\alpha}{\pi \mu(y_0)} \int_t^\infty \int_{-e^{(y+y_0)/2}}^{e^{(y+y_0)/2}} e^{-\alpha y}
		\dd x \dd y
		= \frac{\nu\alpha}{\pi \mu(y_0)} \int_t^\infty 2 e^{(y+y_0)/2} \cdot e^{-\alpha y} \dd y \\
	&= \frac{2 \nu\alpha e^{y_0/2} }{\pi \xi e^{y_0/2} (\alpha-\frac{1}{2}) }  \cdot e^{(\frac{1}{2}-\alpha)t}
		= e^{(\frac{1}{2}-\alpha)t},
\end{align*}
using that $\mu(y_0) = \xi e^{y_0/2} = \left(\frac{2\alpha\nu}{\pi(\alpha-\frac12)}\right) e^{y_0/2}$.
Thus, $y_1, y_2$ are exponentially distributed with parameter $\alpha-\frac12$.
Now note that, for each $t>0$, the probability density $f \cdot 1_{\BallPo{(0,y_0)}} / \mu(y_0)$
is constant on $[-e^{(t+y_0)/2}, e^{(t+y_0)/2}] \times \{t\}$ and it is 
vanishes on $(-\infty, -e^{(t+y_0)/2}) \times \{t\} \cup (e^{(t+y_0)/2},\infty)\times\{t\}$.
